---
title: "Reproduced analysis of FARS data"
author: "Marialexia Alfano"
date: "November 4, 2017"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(knitr)

load("../data/clean_fars.RData")
```

#Results 

##Percentages of drivers testing positive by drug type, sex, and year group

```{r}
clean_fars %>% 
  mutate(year_cat = cut(year, breaks = c(1999, 2002, 2006, 2010),
                        labels = c("1999-2002","2003-2006",
                                   "2007-2010"),
                        include.lowest = TRUE, right = TRUE)) %>% 
  filter(!is.na(sex)) %>% 
  group_by(drug_type, sex, year_cat) %>% 
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test/n_non_missing, 1)) %>% 
  select(drug_type, sex, year_cat, perc_positive) %>% 
  unite(sex_year_cat, sex, year_cat) %>% 
  spread(sex_year_cat, perc_positive) %>% 
  knitr::kable(col.names = c("Drug Type", "F 1999-2002",
                             "F 2003-2006", "F 2007-2010",
                             "M 1999-2002", "M 2003-2006",
                             "M 2007-2010"))
```

##Figure 1: Prevalence of nonalcohol drugs in fatally injured drivers by year and age group 

```{r}
#adjusting data for figure one 
figure_one <- clean_fars %>% 
  filter(drug_type != "Alcohol") %>%
  filter(!is.na(agecat)) %>%
  group_by(unique_id, year, agecat) %>% 
  summarize(positive_tests = sum(positive_for_drug, na.rm = TRUE), positive = any(positive_tests > 0)) %>% 
  ungroup() %>% 
  group_by(year, agecat) %>% 
  summarize(total_tests = length(positive), positive = sum(positive), percent_positive = round(100 * positive/total_tests))
```

```{r}
#plot for figure one
plot_figure_one <- figure_one %>%
  ggplot(aes(x = year, y = percent_positive, position = agecat))+
  geom_point(aes(shape = agecat))+
  geom_line()+
  theme_few()+
  labs(x = "Year", y = "Positive for Nonalcohol Drugs, %", shape = "Age")
  
plot_figure_one
```

##Figure 2: Prevalence of nonalcohol drugs in fatally injured drivers by year and drug type

```{r}
#adjusting data for figure two
figure_two <- clean_fars %>% 
  filter(drug_type != "Alcohol") %>% 
  filter(!is.na(positive_for_drug)) %>% 
  group_by(drug_type, year) %>% 
  summarize(n_not_NA = sum(!is.na(positive_for_drug)),
            positive_tests = sum(positive_for_drug, na.rm = TRUE),
            percent_positive = round(100 * positive_tests/n_not_NA))
```

```{r}
#plot for figure two
plot_figure_two <- figure_two %>% 
  ggplot(aes(x = year, y = percent_positive, position = drug_type))+
  geom_point(aes(shape = drug_type))+
  geom_line()+
  theme_few()+
  scale_y_continuous(breaks = c(4, 8, 12))+
  labs(x = "Year", y = "Positive for Drugs, %", shape = "Drug type")

plot_figure_two
```

##Figure 3: Prevalence of cannabinoid drugs in fatally injured drivers by year and age group

```{r}
#adjusting data for figure three
figure_three <- clean_fars %>%
  filter(drug_type == "Cannabinoid") %>% 
  filter(!is.na(agecat)) %>% 
  group_by(agecat, year) %>% 
  summarize(n_not_NA = sum(!is.na(positive_for_drug)),
            positive_tests = sum(positive_for_drug, na.rm = TRUE),
            percent_positive = round(100 * positive_tests/n_not_NA))
```

```{r}
#plot for figure three
plot_figure_three <- figure_three %>% 
  ggplot(aes(x = year, y = percent_positive, position = agecat))+
  geom_line()+
  geom_point(aes(shape = agecat))+
  theme_few()+
  labs(x = "Year", y = "Positive for Cannabinoid, %", shape = "Age")

plot_figure_three
```

##Prevalence of drugs in fatally injured drivers for 1999 and 2010 by drug type


##Statistics for testing for trend in prevalence of drugs over study years by drug type using Cochran-Armitage trend test
```{r eval = FALSE}
drug_list <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
"Stimulant", "Cannabinoid", "Other")
drug_trend_tests_ca <- lapply(drug_list, test_trend_ca)
drug_trend_tests_ca <- dplyr::bind_rows(drug_trend_tests_ca) %>%
dplyr::mutate(drug = drug_list) %>%
dplyr::select(drug, Z, p.value)
drug_trend_tests_ca %>% knitr::kable()
```

##Statistics for testing for trend in prevalence of drugs over study years by drug type using Wald test of logistic regression coefficient for "year"
```{r eval = FALSE}
drug_list <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
"Stimulant", "Cannabinoid", "Other")
drug_trend_tests_log_reg <- lapply(drug_list, test_trend_log_reg)
drug_trend_tests_log_reg <- dplyr::bind_rows(drug_trend_tests_log_reg) %>%
dplyr::mutate(drug = drug_list) %>%
dplyr::select(drug, Z, p.value)
drug_trend_tests_log_reg %>% knitr::kable()
```

